Comment: Reasoning about Interference Between Units

### 2014-11-24 paper/introduction.tex
Editing

\citet{bowers2013sutva} showed that one could use Fisher's
randomization-based hypothesis testing framework to assess counterfactual
causal models of treatment propagation and spillover across social networks.
That is, they showed that one could make
statements such as: ``Our design provides strong evidence against the idea
that the treatment spilled over by more than 5 points given this model of
treatment propagation.'' This research note shows that the test statistic that
they used --- the Kolmogorov-Smirnov test statistic  \citep[\S
5.4]{MylesHollander1999a} --- could be improved substantially. We demonstrate that the sum of squared residuals from a linear
model in which aspects of the network play a role is a preferable test
statistic for the model and network and design analyzed by
\citet{bowers2013sutva} (henceforth BFP). The point of this short article is to improve the
use of the ``reasoning about interference'' approach and also to highlight
the methodological and statistical questions raised by that article. We do
not aim for general answers to those questions here and now, but we do hope
that this piece stimulates further, and  deeper, work on test statistics and
sharp hypotheses.

\section{Background}

We can define a counterfactual causal effect for a person $i$ as a difference
between the outcome, $y_i$, person $i$ would have expressed
after a set of treatments, collected in the vector $\bz$, had been assigned to some subset of a social network,
$y_{i,\bz}$ compared to the outcome that person $i$ would have expressed
under a different treatment vector. If the experiment had an effect on person
$i$, then her outcome would differ under different realizations of the
experimental treatment $y_{i,\bz} \ne y_{i,\bz'}$. The fundamental problem of
causal inference (a phrase famously coined in \cite{holland:1986a}) reminds
us that we can never see both states of the world: we only observe the
outcome from person $i$ under one treatment assignment, either $\bz$ or
$\bz'$ not both.

R.A. Fisher's approach to design-based statistical inference  \citep[Chap
2]{fisher:1935} begins with the idea that we do not learn directly about the
unobserved counterfactuals. Rather, he suggests, we can learn about models of
unobserved counterfactuals --- we can learn how much information we have to
dispel certain claims about those counterfactuals as implausible. This
conceptual move --- sidestepping the fundamental problem of causal inference
via learning about claims made by scientists --- drives hypothesis testing in
general.  BFP build on this insight as developed and
linked to counterfactual causal inference by Paul Rosenbaum
\citep{rosenbaum2010design}. They also say: we do not know how the
counterfactual conditions might have turned out, they are unobserved, but we
can make claims about how treatment might have come to have an effect and we
can assess those claims. For example, the claim used by
BFP is a model in which the effects of treatment
propagating from one person to another die off as the network distance between
nodes increases.\footnote{See the paper itself for details of the example
  model.} They further showed that the strength of evidence against the
specific hypotheses implied by a given model varied with different features of
the design as well as the extent to which the true process by which treatment
given to one unit changed the outcomes of another unit diverged from the
model. Throughout their paper they used the Kolmogorov-Smirnov (KS) test
statistic so that their tests would be sensitive differences in the treatment
and control distributions implied by different hypotheses and not merely
sensitive to differences in one aspect of those distributions (such as the
differences in the mean).\footnote{If the empirical cumulative distribution
  function (ECDF) of the treated units is $F_1$ and the ECDF of the control
  units is $F_0$ then the KS test statistic is $\T(\yu,\bz) = \underset{i =
    1,\ldots,n}{\text{max}}\left[F_1(y_{i,\bzero}) -
    F_0(y_{i,\bzero})\right]$, where $F(x)=(1/n)\sum_{i=1}^n I(x_i \le x)$
  records the proportion of the distribution of $x$ at or below $x_i$
  \citep[\S 5.4]{MylesHollander1999a}.}

Notice that the test statistic choice matters: the engine of statistical
inference involves summarizing information against the hypothesized claim, yet
different ways to summarize information might be more or less sensitive to
substantively meaningful differences. A simple test of the
sharp null can be executed with power that varies as a function of the design
of the study (proportion treated, blocking structure, etc..), characteristics
of the outcome, and the way that a test statistic summarizes the outcome. Yet,
when the models are complex,  BFP showed that sometimes a
test will have (1) no power to address the model at all such that all hypothesized parameters
would receive the same implausibility assessment; (2) or might reject all such
hypothesized parameters. Thus, although in theory one may assess sharp
hypotheses about multiparameter settings, in practice one may not learn much
from such tests --- or, in the absence of simulation studies assessing the
model and test statistic, one may be mislead.

How should one choose a test statistic to assess relationships between complex
models of treatment propagation and data? General advice for test statistics
is that they should be powerful against relevant alternatives (find canonical
cite). \citet[\S 2.4.4]{rosenbaum:2002} provides more specific advice about
large sample performance of certain classes of test statistics and BFP repeat
his general advice: `` Select a test statistic T that will be small when the
treated and control distributions in the adjusted data \ldots are
similar, and large when the distributions diverge.'' \citep[p.
107]{bowers2013sutva}. They should have noted, instead, that when models
involve more than one parameter, such as the models of treatment propagation across
networks, test statistics should be increasing in \emph{all} of the parameters
(for example, in both parameters capturing spillover and parameters capturing
direct experience with the treatment). Here, we also show a benefit from
taking into account the network structure in the test statistic as well as in
the model.

We demonstrate two results, each showing how effect increasing test statistics,
as defined by BFP, may provide little to no information about the spillover
parameter $\tau_2$.  Let us first consider the case when the test statistic is
the absolute mean difference between hypothesized uniformity trial outcomes
for treated and control individuals: $\T(\widetilde y_{ij}(\0,\Z), \Z') =
\left\vert \mu_1 ( \widetilde y_{ij}(\0,\Z), \Z' ) - \mu_0 ( \widetilde
  y_{ij}(\0,\Z), \Z' ) \right\vert.  $ Fix a realization of $\Z = \z$, and
further assume noncollinearity: $ \z \notin \{\z^I, (1-\z^I)\}$. We can see
that, for any hypothesized $\widetilde\tau_2$, there always exists a
$\widetilde\tau_1$ such that the $p$-value is 1. Expanding the test statistic,

\begin{flalign}
  \T(\widetilde y_{ij}(\0,\z), \z) = & \left\vert \mu_1 (y_{ij}(\z) - \widetilde\tau_1 - \tau_2 z_{i(-j)}, \z) - \mu_0(\widetilde y_{ij}(\z) -
  \widetilde\tau_2 z_{i(-j)} , \z) \right\vert \\
= & \left\vert \mu_1 (y_{ij}(\z)
  - \widetilde\tau_2 z_{i(-j)}, \z)- \mu_0 (\widetilde y_{ij}(\z) -
  \widetilde\tau_2 z_{i(-j)}, \z)- \widetilde\tau_1 \right\vert.
\end{flalign}

Thus,
regardless of the value of $\widetilde\tau_2$, we can always find a
$\widetilde\tau_1$ that sets $\T(\widetilde y_{ij}(\0,\z), \z) = 0$. Given
that $\T(\widetilde y_{ij}(\0,\z), \Z')$ can only take values in
$\mathbb{R}^+$, then $p=1$, as any other possible randomization $\Z'$ will
yield a value of $\T(\widetilde y_{ij}(\0,\z), \Z') \geq 0$. It follows that,
given this test, we cannot rule out any value of $\tau_2$ as being
implausible.\footnote{In fact, this result does not depend on example in the
  setting, and holds for any randomization scheme coupled with a linear causal
  model of the form \eqref{causal}.}

Second, I show that, given the running example, a test based on any effect
increasing statistic is inconsistent against a radically incorrect hypothesis
about $\tau_2$. Let us assume that $\tau_1$ is properly hypothesized at its
true value, and that the hypothesized value of $\tau_2$ ($\widetilde \tau_2$)
is radically incorrect: $\vert \widetilde\tau_2 - \tau_2 \vert \gg
\max(y_{ij}(\0)) - \min(y_{ij}(\0))$. Denote $\beta \equiv \tau_2 -
\widetilde\tau_2$, and assume $\vert \beta \vert < \infty$.  Under the
assumption that $\tau_2 = \widetilde \tau_2$: $\widetilde y_{ij}(\0,\Z) =
y_{ij}(\0) + \beta Z_{i(-j)}$.  Assume a sequence of nested finite populations
$\{U_k\} = U_1, U_2,...$, growing in $k$ households, so that $U_k$ consists of
$N_k = 2k$ individuals, each with an associated random assignment $\Z_k$
\citep{isaki1982survey}. (For notational parsimony, I drop explicit dependence
on the subscript $k$.)  Given an incorrect hypothesis, a necessary, but not
sufficient, condition for a test's consistency is that $\lim_{k \rightarrow
\infty} \Pr(p > \alpha) < \epsilon$, for any $\alpha,\epsilon > 0$.

Let an effect increasing test statistic $\T(\widetilde y_{ij}(\0,\Z),\Z')$ be
some measure of the absolute divergence in empirical distributions of
$\widetilde y_{ij}(\0,\Z)$ for treated and control individuals under the
treatment vector $\Z'$.  Since distributional differences in $y_{ij}(\0)$ are
small relative to shifts induced by differences in proportions of treated
housemates,  $\T(\widetilde y_{ij}(\0,\Z),\Z')$ will be increasing in $
\left\vert \mu_1(Z_{i(-j)},\Z') -   \mu_0(Z_{i(-j)},\Z') \right\vert$. , For
a realization $\Z = \z$, the hypothesized randomization distribution of
$\mu_1(z_{i(-j)},\Z') - \mu_0(z_{i(-j)},\Z')$ is asymptotically normal with
mean $0$ and variance $V_{x(\z)} = 4 x(\z) \left[1-x(\z)\right] / N$, where
$x(\z) \equiv \sum_{ij\in U}z _{ij} /N$, if $0 < x(\z) <  1$.  (by the
Lindeberg CLT, Taylor linearization and Slutsky's theorem).  Thus the
$p$-value associated with randomization $\z$, $p = 2
\Phi(-|\mu_1(z_{i(-j)},\z) - \mu_0(z_{i(-j)},\z)|/V_{x(\z)}^{1/2})$. It
follows that $|\mu_1(z_{i(-j)},\z) - \mu_0(z_{i(-j)},\z)\vert <
-\Phi^{-1}(\alpha/2) V_{x(\z)}^{1/2} \Leftrightarrow p > \alpha$.  By similar
calculations at the household level, the (actual, not hypothesized)
randomization distribution of $\mu_1(Z_{i(-j)},\Z) - \mu_0(Z_{i(-j)},\Z)$ is
asymptotically mean $0$ with variance $2 V_{1/2} =  2/N$.\footnote{This result
  follows from reframing the problem as randomly assigning households into 4
  conditions \{00,01,10,11\} with equal probability, then linearizing
  $\mu_1(Z_{i(-j)},\Z) - \mu_0(Z_{i(-j)},\Z)$.} By the LLN ,  $\plim_{k
\rightarrow \infty} x = 1/2$, and so, by the and continuous mapping theorem,
$\lim_{k \rightarrow \infty} \Pr(\vert NV_{x(\Z)} - NV_{1/2}\vert >
\epsilon_3) = 0$.  Then, by Chebyshev's inequality, $\lim_{k \rightarrow
\infty}  \Pr(p > \alpha) > 1- 2\left[\Phi^{-1}(\alpha/2)\right]^{-2}$,
$\epsilon = 1- 2\left[\Phi^{-1}(\alpha/2)\right]^{-2}$ proportion of possible
$\Z \in \Omega$ have $p > \alpha$,
for any $0 < \alpha < 2\Phi(-\sqrt{2})$.  Tests based on effect increasing
statistics are therefore not generally consistent.

>>>>

\citet{bowers2013sutva} showed that one could use Fisher's randomization-based
hypothesis testing framework to assess counterfactual causal models of
treatment propagation and spillover across social networks.  This research
note improves the statistical infernce presented in \citet{bowers2013sutva}
(henceforth BFP) by substituting a test statistic based on a sum of squared
residuals and incorporating information about the fixed network for the simple
the Kolmogorov-Smirnov test statistic  \citep[\S 5.4]{MylesHollander1999a}
that they used. The point of this short article is to incrementally improve
the use of BFP's ``reasoning about interference'' approach. We do not offer
results about test statistics for multi-parameter causal models on social
networks here, but instead hope that this research note stimulates further,
and deeper, work on test statistics and sharp hypotheses.

\section{Background on randomization based statistical inference for questions
about causal effects from social network experiments}

In a randomized experiment with $n=4$ subjects connected via a fixed network,
subject $i=1$ might respond differently to different ways that treatment is
assigned to the whole network. When the treatment assignment vector provides
treatment to persons 2 and 3, $\bz=\{0,1,1,0\}$ she might act one way,
$y_{i=1,\bz=\{0,1,1,0\}}$ and when treatment is assigned to persons 3 and 4,
$\bz=\{0,0,1,1\}$ she might act another way, $y_{i=1,\bz=\{0,0,1,1\}}$ That
is, if the experiment had a causal effect on person $i$, then her outcome
would differ under different realizations of the experimental treatment
$y_{i,\bz} \ne y_{i,\bz'}$. The fundamental problem of causal inference
reminds us that we can never see both states of the world: we only observe the
outcome from person $i$ under one treatment assignment vector, either $\bz$ or
some $\bz'$ not both  \cite{holland:1986a}.

R.A. Fisher's approach to design-based statistical inference  \citep[Chap
2]{fisher:1935} as developed by Paul Rosenbaum  \citep{rosenbaum2010design}
begins with the premise of the fundamental problem of causal inference. Rather
than imagine that we could learn directly about the unobservable
counterfactual comparisons of outcomes under different treatment vectors,
Fisher would suggest that we focus on learning about how \emph{models} of
those counterfactual outcomes relate to what we observe. Although we do not
know how person $i$ would have acted under all possible experimental
scenarios, we can learn how much information we have to dispel certain claims
or hypotheses.  This conceptual move --- sidestepping the fundamental problem
of causal inference via learning about claims made by scientists --- drives
hypothesis testing in general.  BFP build on this insight by showing that
models of counterfactual effects can involve statements about how treatment
given to one node in a social network can come to have an effect on other
nodes. Their example model allows the effects of treatment to die off as the
network distance between nodes increases.\footnote{See the paper itself for
  details of the example model.} They further showed that the strength of
evidence against the specific hypotheses implied by a given model varied with
different features of the research design as well as the extent to which the
true causal process diverged from the model. Since their simulated experiment
involved only two treatments, the only observations available to evaluate the
model were comparisons of the assigned to treatment group and the assigned to
control group. Since their model could imply not only shifts in the mean of
the observed treatment versus control outcome distributions, but changes in
shape of those distributions, they used the Kolmogorov-Smirnov (KS) test
statistic so that their tests would be sensitive differences in the treatment
and control distributions implied by different hypotheses and not merely
sensitive to differences in one aspect of those distributions (such as the
differences in the mean).\footnote{If the empirical cumulative distribution
  function (ECDF) of the treated units is $F_1$ and the ECDF of the control
  units is $F_0$ then the KS test statistic is $\T(\yu,\bz) = \underset{i =
    1,\ldots,n}{\text{max}}\left[F_1(y_{i,\bzero}) -
    F_0(y_{i,\bzero})\right]$, where $F(x)=(1/n)\sum_{i=1}^n I(x_i \le x)$
  records the proportion of the distribution of $x$ at or below $x_i$
  \citep[\S 5.4]{MylesHollander1999a}.} So, in broad outline, the BFP approach
involves (1) the articulation of a model for how a treatment assignment vector
can change outcomes for all subjects in the experiment (holding the network
fixed) and (2) using function of comparing actually treated and control
observations to tell us whether such a model is implausible (codified as a low
$p$-value)  or whether we have too little information available from the data
and design about a the model (codified as a high $p$-value). This is classic
hypothesis testing.

Notice that the test statistic choice matters in this process: the engine of
statistical inference involves summarizing information against the
hypothesized claim, yet different ways to summarize information might be more
or less sensitive to substantively meaningful differences. The statistical
power of a simple test of the sharp null hypothesis of no effects will vary as
a function of the design of the study (proportion treated, blocking structure,
etc.), characteristics of the outcome (continuous, binary, skewed, extreme
points, etc.), and the way that a test statistic summarizes the outcome (does
it compare means, standard deviations, medians, qqplots, etc.). In general,
test statistics should be powerful against relevant alternatives (find
canonical cite). \citet[\S 2.4.4]{rosenbaum:2002} provides more specific
advice about large sample performance of certain classes of test statistics
and BFP repeat his general advice: `` Select a test statistic T that will be
small when the treated and control distributions in the adjusted data \ldots
are similar, and large when the distributions diverge.'' \citet[Proposition 4
and 5, \S 2.9]{rosenbaum:2002} presents results proving that test statistics
with this property (``effect increasing'' test statistics), produces an
unbiased test of the hypothesis of no effects or positive effects when the
positive effects involve one parameter.  Yet, when the models are complex and
may involve increasing effects in the direction of one parameter and
non-linear effects in the direction of another parameter, BFP showed that
sometimes a KS-test will have (1) no power to address the model at all such
that all hypothesized parameters would receive the same implausibility
assessment; (2) or might reject all such hypothesized parameters. Thus,
although in theory one may assess sharp hypotheses about multiparameter
settings, in practice one may not learn much from such tests --- or, in the
absence of simulation studies assessing the model and test statistic, one may
be misled. BFP thus recommended simulation studies of the operating
characteristics of tests as a piece of their workflow --- because  the theory
justifying simple one-dimensional effect increasing test statistics clearly
did not cover multi-parameter situations like those easily arising from social
network experiments.

### 2014-11-24 paper/paper.tex
Starting writing
Editing

\author{Peter Aronow \thanks{Assistant Professor, Dept of Political Science,
    Yale University} \and Jake Bowers \thanks{Associate Professor, Dept of Political
    Science and Statistics, University of Illinois @ Urbana-Champaign
    (\texttt{jwbowers@illinois.edu}). \textit{Acknowledgements:} Part
    of this work funded by NSF Grants SES-0753168 and SES-0753164
    (joint with Ben Hansen, Dept of Statistics, University of
    Michigan).} \and Mark Fredrickson \thanks{Ph.D. Student in Statistics and Political Science, University of Illinois @ Urbana-Champaign.}}

@@

% this appendix is the response to reviewers. TODO: remove for released
% version
% \clearpage
% \setlength{\parindent}{0pt}
% \addtolength{\parskip}{.5\baselineskip}
% \pagenumbering{arabic} %\setcounter{page}{1}
% \setcounter{mycounter}{0}
% \setcounter{secnumdepth}{0}
% %\let\osection\section
% %\renewcommand{\section}{\stepcounter{mycounter}\osection}
% \renewcommand\thesection{\arabic{mycounter}}
% \input{response}
% 
\end{singlespacing}
%\bibliography{/Users/jwbowers/Documents/BIB/trunk/big}
\end{document}

>>>>

\author{Peter Aronow \thanks{Assistant Professor, Dept of Political Science,
    Yale University \textit{Acknowledgements:} } \and Jake Bowers \thanks{Associate Professor, Dept of Political
    Science and Statistics, University of Illinois @ Urbana-Champaign
    (\texttt{jwbowers@illinois.edu}).} \and Mark Fredrickson \thanks{Ph.D. Student in Statistics and Political Science, University of Illinois @ Urbana-Champaign.}}

@@

% appendix 
% \clearpage
% \setlength{\parindent}{0pt}
% \addtolength{\parskip}{.5\baselineskip}
% \pagenumbering{arabic} %\setcounter{page}{1}
% \setcounter{mycounter}{0}
% \setcounter{secnumdepth}{0}
% %\let\osection\section
% %\renewcommand{\section}{\stepcounter{mycounter}\osection}
% \renewcommand\thesection{\arabic{mycounter}}
% \input{response}
%
\end{singlespacing}
%\bibliography{/Users/jwbowers/Documents/BIB/trunk/big}
\end{document}
